---
title: "Data Vizualization in R I"
author: "Abel Torres Espin"
date: "Nov 2022"
output: 
    html_document:
      toc: true
      toc_float: true
      theme: readable
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, error = FALSE, warning = FALSE)
```

# Libraries

```{r}
# install.packages('gridExtra')
# install.packages('patchwork')
# install.packages('ggplot2')

library(ggplot2) #the implementation of the grammar of graphics
library(gridExtra) #a ggplot2 extension for layouts
library(patchwork) #a ggplot2 extension for layouts
```

# Data 
```{r}
# ?ChickWeight #Check the documentation

df<-ChickWeight #easier name for the data contained in a dataframe

head(df, 10) #Quick look of the dataframe content
str(df) #Quick look of the variables

summary(df) #Quick summary of the variables
```

# Grammar of Graphics in R: ggplot2 

```{r, eval=FALSE}
#ggplot2 package
?ggplot2
```

## ggplot2 basic structure, the ggplot function

```{r, eval=FALSE}
?ggplot()
```

### Data and aesthetics

```{r}
ggplot(data=df, aes(x=Time, y=weight))
```

### Geometry

```{r, eval=FALSE}
ggplot(df, aes(Time, weight))+
  geom_point()
```

## Statistic

```{r}
ggplot(df, aes(Time, weight))+
  geom_point()+
  stat_smooth(method = lm)
```

## Theme

```{r}
ggplot(df, aes(Time, weight))+
  geom_point()+
  stat_smooth(method = lm)+
  theme_classic()+
  xlab('Time (days)')+
  ylab('Weight (g)')

ggplot(df, aes(Time, weight))+
  geom_point(fill='green',color='black', shape=23, size=2, alpha=0.2)+
  stat_smooth(method = lm, color='red')+
  xlab('Time (days)')+
  ylab('Weight (g)')+
  theme(axis.text = element_blank(), 
        panel.background = element_rect(fill = "purple"))
```

## More ggplot2
### Layers

```{r}
ggplot(df, aes(Time, weight))+
  geom_point(fill='grey',color='black', shape=21, size=2)+
  stat_smooth(method = lm, color='red')+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')

ggplot(df, aes(Time, weight))+
  geom_boxplot()+
  geom_point(fill='grey',color='black', shape=21, size=2)+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')
```

Time is numeric, so ggplot interpret it as continous. If we want a boxplot for each timepoint we have to transform it to a factor

```{r}
ggplot(df, aes(factor(Time), weight))+
  geom_boxplot()+
  geom_jitter(fill='grey',color='black', shape=21,width=1,
              size=2)+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')

ggplot(df, aes(factor(Time), weight))+
  theme_minimal()+
  geom_boxplot()+
  geom_point(fill='grey',color='black', width=0.5,
             shape=21, size=2)+
  
  xlab('Time (days)')+
  ylab('Weight (g)')
```

### Grouping

```{r}
ggplot(df, aes(Time, weight))+
  geom_point()

ggplot(df, aes(Time, weight, group=Diet))+
  geom_point()

ggplot(df, aes(Time, weight, group=Diet))+
  geom_point()+
  stat_smooth(method = lm)

ggplot(df, aes(Time, weight, color=Diet))+
  geom_point()+
  stat_smooth(method = lm)

ggplot(df, aes(Time, weight, size=weight, color=Diet))+
  geom_point()+
  stat_smooth(method = lm)

ggplot(df, aes(Time, weight, shape=Diet, size=weight, color=Diet))+
  geom_point()+
  stat_smooth(method = lm)
```

### Faceting

```{r}
ggplot(df, aes(Time, weight, color=Diet))+
  geom_point()+
  stat_smooth(method = lm, color='blue')+
  facet_grid(~Diet)

ggplot(df, aes(Diet, weight, color=Diet))+
  geom_boxplot()+
  stat_smooth(method = lm, color='blue')+
  facet_grid(~Time)
```

Next one is computationally more intensive

```{r}
ggplot(df, aes(Chick, weight, color=Diet))+
  geom_boxplot()+
  stat_smooth(method = lm, color='blue')+
  facet_grid(Diet~Time)
```

### More on statistics

```{r}
ggplot(df, aes(Time, weight))+
  geom_point(fill='grey',color='black', shape=21, size=2)+
  stat_smooth(method = lm, color='red')+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')

ggplot(df, aes(Time, weight, group=Diet, fill=Diet, color=Diet))+
  geom_point(size=2)+
  geom_line()+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')+
  facet_grid(~Diet)

ggplot(df, aes(Time, weight, group=Diet, fill=Diet, color=Diet))+
  # geom_point(size=2)+
  stat_summary(fun = 'mean', geom='line')+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')

ggplot(df, aes(Time, weight, group=Diet, fill=Diet, color=Diet))+
  # geom_point(size=1, alpha=0.4)+
  stat_summary(fun = 'mean', geom='point', size=2)+
  stat_summary(fun = 'mean', geom='line')+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')

ggplot(df, aes(Time, weight, group=Diet, fill=Diet, color=Diet))+
  # geom_point(size=1, alpha=0.4)+
  stat_summary(fun = mean, geom='line')+
  stat_summary(fun = mean, geom='point', size=2)+
  stat_summary(fun.data =mean_se, geom='errorbar')+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')

ggplot(df, aes(Time, weight, group=Diet, fill=Diet, color=Diet))+
  stat_summary(fun = 'mean', geom='line')+
  stat_summary(fun = 'mean', geom='point', size=2)+
  stat_summary(fun.data =mean_cl_normal, geom='errorbar', alpha=0.2, aes(color=NULL))+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')

ggplot(df, aes(factor(Time), weight, group=Diet, fill=Diet, color=Diet))+
  stat_summary(fun = 'mean', geom='bar', 
               position = position_dodge())+
  stat_summary(fun.data =mean_se, geom='errorbar', position=position_dodge())+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')
```


### Colors and shapes
```{r}
colors<-c('green','orange', 'firebrick', 'steelblue')
shapes<-c(15,17,16,5)

ggplot(df, aes(Time, weight, group=Diet, color=Diet, shape=Diet))+
  stat_summary(fun.y = 'mean', geom='line')+
  stat_summary(fun.y = 'mean', geom='point', size=2)+
  stat_summary(fun.data =mean_se, geom='errorbar')+
  scale_color_manual(values=colors)+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')

ggplot(df, aes(Time, weight, group=Diet, color=Diet, shape=Diet))+
  stat_summary(fun.y = 'mean', geom='line')+
  stat_summary(fun.y = 'mean', geom='point', size=2)+
  stat_summary(fun.data =mean_se, geom='errorbar')+
  scale_color_manual(values=colors)+
  scale_shape_manual(values=shapes)+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')
```

## ggplot as object

ggplot does not generates an image directly as the plot function does. It creates and object. If we do not assign that object to a variable ggplot function will automatically call the plot function and plot the graph.

```{r}
myplot<-ggplot(df, aes(Time, weight, group=Diet, color=Diet, shape=Diet))+
  stat_summary(fun.y = 'mean', geom='line')+
  stat_summary(fun.y = 'mean', geom='point', size=2)+
  stat_summary(fun.data =mean_se, geom='errorbar')+
  scale_color_manual(values=colors)+
  scale_shape_manual(values=shapes)+
  theme_minimal()+
  xlab('Time (days)')+
  ylab('Weight (g)')
```

```{r}
myplot
```

```{r}
myplot+theme_bw()+ggtitle('title')
```


## Annotations
### Adding lines

```{r}
myplot+
  geom_vline(xintercept = 10, color='blue',linetype='solid')+
  geom_hline(yintercept = 100, color='red', size=2)
```

### Some free text

```{r}
myplot+
  annotate(geom='text',label='this is some random text', x = 10,
           y=250)
```

### Mapping some text

```{r}
myplot+
  aes(label=Chick)+
  geom_text()
```

## Saving plots

```{r}
myplot
ggsave(plot = myplot, filename = 'plots/myplot.jpeg', dpi = 300)
```

```{r}
pdf(file = 'myplot.pdf')
myplot
dev.off()
```

```{r}
png(file='myplot.png', res = 300)
myplot
dev.off()
```

other options are tiff, jpeg, svg...

# Figures: Bringing it all together

```{r}
df$height<-5+0.07*df$weight+0.01*df$weight^2+
  rnorm(n=length(df$weight), sd=50)

pplot<-ggplot(df, aes(color=Diet, shape=Diet, group = Diet))+
  scale_color_manual(values=colors)+
  scale_shape_manual(values=shapes)+
  theme(axis.line = element_line(color='black'),
        panel.background = element_blank(), 
        panel.grid = element_line(color='grey90'),
        text = element_text(size=12))
# pplot
```

```{r}
p1<-pplot+aes(x=Time,y=weight)+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5)+
  ylab('Weight (g)')+xlab('Time (days)')+
  theme(legend.position = 'none')
p1
```

```{r}
p2<-pplot+aes(x=Time,y=height)+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5)+
  ylab('Height (mm)')+xlab('Time (days)')+
  theme(legend.position = 'none')
p2

p2<-p1+aes(y = height)+ylab('Height (mm)')

```

```{r}
p3<-pplot+aes(x=height, y=weight)+
  geom_point(alpha=0.2)+
  geom_smooth(se = F)+
  facet_grid(~Diet)+
  xlab('Height (mm)')+ylab('Weight (g)')

p3
```

## gridExtra

[How to use gridExtra](https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html)

```{r}
figure1<-grid.arrange(p1, p2, p3, nrow=2, ncol=2,
                      layout_matrix=rbind(c(1,2),c(3,3)))

# plot(figure1)

ggsave(plot = figure1,filename = 'figure1.png')
```

## patchwork

[patchwork website](https://patchwork.data-imaginist.com/)

```{r}
(p1+p2)/p3
```

```{r}
(p1+p2)/p3+plot_annotation(tag_levels = "a")
```

## The great of coding plots

```{r}
colors <-c("black", "red", "blue", "purple")

pplot<-ggplot(df, aes(color=Diet, shape=Diet, group = Diet))+
  scale_color_manual(values=colors)+
  scale_shape_manual(values=shapes)+
  theme(axis.line = element_line(color='black'),
        panel.background = element_blank(),
        panel.grid = element_line(color='grey90'),
        text = element_text(size=12))

p1<-pplot+aes(x=Time,y=weight)+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5)+
  ylab('Weight (g)')+xlab('Time (days)')+
  theme(legend.position = 'none')

p2<-pplot+aes(x=Time,y=height)+
  stat_summary(fun = mean, geom = "line")+
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5)+
  ylab('Height (mm)')+xlab('Time (days)')+
  theme(legend.position = 'none')

p3<-pplot+aes(x=height, y=weight)+
  geom_point(alpha=0.2)+
  geom_smooth(se = F)+
  facet_grid(~Diet)+
  xlab('Height (mm)')+ylab('Weight (g)')

(p1+p2)/p3+plot_annotation(tag_levels = "a")
```




